Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
update

We benchmark our algorithms by computing various independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performacne.

@@

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of matrix multiplication to multi-dimensional tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ the matrix multiplication between two matrices $A$ and $B$ can be represented in Einstein's notation as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard notation of Einstein's summation or tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is a vertex and two tensors sharing the same label are connected.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
%In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of these generalized tensor networks is hypergraphs, in which an edge (label) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, which is a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

This motivates us to think about what is the most general element type allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
The requirement of being commutative is for the tensor contraction result to be independent of the contraction order.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings while keeping the tensor network generic~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

@@

\section{Independence polynomial} \label{sec:indpoly}
% \subsection{Independence polynomial}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute the independence polynomial of a graph $G = (V, E)$, we reinterpret this problem as a tensor network contraction problem.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evaluate \Eq{eq:idp}, directly summing up the products is computationally inefficient.
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor elements.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{\text{tw}(G)})$~\cite{Markov2008}.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

\section{Benchmarks and case studies}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small tree width that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

Figure (a) shows the time and space complexity (without slicing) of tensor network contraction for variety of graph sizes, where the space complexity is the same as the tree width of the problem graph.
It is obtained with tensor network contraction order optimization algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used to fit the computation into a 32GB memory, the target tree width of slicing is $27$.
One can see all the computing times in figure (b), (c) and (d) have a strong correlation with the tree width.
%The computations shown in \Fig{fig:benchmark} include independent set counting, finding maximum independent set size (panel b),
%computing independence polynomial (panel c) and find one independent set and enumerating independent sets (panel d).
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions, hence 
are much faster comparing to non-BLAS methods in the same category.
GPU computes much faster than CPU in all cases when the problem scale is large enough so that the actual computing time is comparable or larger than the launching time of CUDA kernels.
Most algebras can be computed on GPU, except those requiring dynamic sized structures, i.e. \texttt{PN} and \texttt{SN}.
In figure (c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
however, it maybe suffer from the round off errors. The finite field (GF$(p)$) approach is the only method that does not incur round-off errors and can run on a GPU.
In figure (d), one can see the technique to bound the enumeration space improves the performance for more than one order in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a 32GB memory device.

@@

\subsubsection{Number of independent sets and hard-square entropy constant}
In this examples, the types of graphs we used are shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The graphs include: the square grid graphs, denoted as SG; the square grid graphs with a filling factor $p$, denoted as SG$(p)$, where $\lfloor pL^{2} \rceil$ square grids are occupied with vertices;
the King's graphs, denoted as K; the King's graphs with a filling factor $p$, denoted as K$(p)$. 
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequences for $L=38$ and $L=39$, which is not known before to our knowledge.
In the computation, we have used the finite field algebra trick for contracting arbitrarily high precision integer tensor networks.
We also computed the quantity $F(L,L)^{1/\lfloor pL^2\rceil}$ as a function of grid size $L$ for these lattices,
where $F(L,L)$ is the number of independent sets of a given grid size.
The results are shown in \Fig{fig:hardsquare}.
Our result for $p=1$ matches those in the literature, and the lattices with hole has much larger average value,
which has a phsical meaning of the hard lattices with holes have much larger mean entropy than those without holes at the high temperature limit.

>>>>

We benchmark our algorithms in \Sec{sec:benchmark} by computing various independent set properties of certain sparse graphs on central processing units (CPUs) and graphics processing units (GPUs) and show its good performacne.

@@

\section{Tensor networks}
Tensor network~\cite{Cirac2020, Orus2014} is also known as einsum, factor graph or sum-product network~\cite{Bishop2006} in different contexts;
it can be viewed as a generalization of binary matrix multiplication to nary tensor contraction.
Einstein's notation is often used to represent a tensor network, e.g.\ it represents the matrix multiplication between two matrices $A$ and $B$ as $C_{ik} = A_{ij}B_{jk}$,
where we use a label in the subscript to represent a degree of freedom.
One can enumerate these degrees of freedom and accumulate the product of tensor elements to the output tensor.
In the standard Einstein's notation for tensor networks in physics, each index appears precisely twice.
Hence a tensor network can be represented as a simple graph,
where a tensor is mapped to a vertex and a label shared by two tensors is mapped to an edge.
In this work, we do not impose such a restriction, so an index can appear an arbitrary number of times. 
%In the main text, we do not have this restriction. An index can appear arbitrary times.
The graphical representation of a generalized tensor network is a hypergraph, in which an edge (a label in subscript) can be shared by an arbitrary number of vertices (tensors).
These two notations are equivalent in representation power because one can easily translate a generalized tensor network to the standard notation by adding $\delta$ tensors, a high dimensional equivalence of the identity matrix.
However, introducing $\delta$ tensors may significantly increase the complexity of the tensor contraction. We illustrate this subtle point in \App{app:tensorbad}.

@@

This motivates us to think about what is the most general element type allowed in a tensor network contraction program.
We find that as long as the algebra of tensor elements forms a commutative semiring, the tensor network contraction result will be valid and be independent of the contraction order.
A commutative semiring is a ring with its multiplication operation being commutative and without an additive inverse.
To define a commutative semiring with the addition operation $\oplus$ and the multiplication operation $\odot$ on a set $R$, the following relations must hold for any arbitrary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b \oplus a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}
We require the algebra being multiplicative commutative because the contraction order optimizer will change the multiplication order of elements, while we wish this contraction order does not change the contraction results.
In the following sections, we show how to compute a number of properties of independent sets by designing tensor element types as specific commutative semirings without changing the tensor network contraction program~\cite{Stepanov2014}.
Table~\ref{tbl:generictypes} summarizes those properties that can be computed by various tensor element types.

@@

\section{Independence polynomial} \label{sec:indpoly}
% \subsection{Independence polynomial}
The independence polynomial is an important graph polynomial that contains the counting information of independent sets. It is defined as
\begin{equation}\label{eq:idpdef}
I(G, x) = \sum_{k=0}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in graph $G=(V,E)$. The total number of independent sets is thus equal to $I(G, 1)$.
To compute this polynomial, we map this expression to a tensor network contraction.
We map a vertex $i\in V$ to a label $s_i \in \{0, 1\}$ of dimension $2$ in a tensor network, where we use $0$ ($1$) to denote a vertex is absent (present) in the set.
For each label $s_i$, we defined a parametrized rank-one vertex tensor $W(x_i)$ as
\begin{equation}
    W(x_i) = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right).
\end{equation}
We use subscripts to index tensor elements, e.g.\ $W(x_i)_0=1$ is the first element associated with $s_i=0$ and $W(x_i)_1=x_i$ is the second element associated with $s_i=1$.
Similarly, on each edge $(u, v)$, we define a matrix $B$ indexed by $s_u$ and $s_v$ as
\begin{equation}
    \qquad \quad 
       B = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right). \label{eq:edgetensor}
\end{equation}
The corresponding tensor network contraction gives
\begin{equation}\label{eq:idp}
    P(G, \{x_1, x_{2}, \ldots,x_{|V|}\}) = \sum\limits_{s_1, s_2, \ldots, s_{|V|} = 0}^{1} \prod\limits_{i=1}^{|V|} W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j},
\end{equation}
where the summation runs over all vertex configurations $\{s_1, s_{2}, \ldots,s_{|V|}\}$ and accumulates the product of tensor elements to the output $P$ (see Example \ref{eg:tensorcontraction} for a concrete example).
The edge tensor element $B_{s_{i}=1, s_{j}=1} = 0$ encodes the independent set constraint, meaning vertex $i$ and $j$ cannot be both in the independent set if they are connected by an edge $(i,j)$.
In the special case of $x_i = x$, the contraction result directly corresponds to the independence polynomial.
The connection can be understood as follows: the product over vertex tensor elements produces a factor $x^k$, where $k=\sum_i s_i$ counts the set size,
and the product over edge tensor elements gives a factor $1$ for a configuration being in an independent set and $0$ otherwise. The summation counts the number of independent sets of size $k$.

%In the implementation of the algorithm, we benefit from recent advances in tensor network contraction,
%where researchers working on quantum circuit simulation evaluate the tensor network by pairwise contracting tensors in a heuristic order~\cite{Gray2021, Pan2021}.
To evaluate \Eq{eq:idp}, directly summing up the $2^{|V|}$ product terms is computationally inefficient.
The standard approach to evaluate a tensor network is to contract two tensors at a time with a certain order utilizing the associativity and commutativity of tensor contraction.
A fully optimized contraction order can reduce the time to $O(2^{\text{tw}(G)})$~\cite{Markov2008}, while requiring a space at the same order of magnitude to store intermediate contraction results.
The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.

@@

\section{Benchmarks and case studies}\label{sec:benchmark}
\subsection{Performance benchmarks}
We run a single thread benchmark on CPU Intel(R) Xeon(R) CPU E5-2686 v4 @ 2.30GHz,
and its CUDA version on a GPU Tesla V100-SXM2 16G.
The results are summarized in Figure~\ref{fig:benchmark}.
The graphs that we use in benchmarks are random three regular graphs,
 a typical type of sparse graphs that has a small tree width that asymptotically smaller than $|V|/6$~\cite{Fomin2006}.

@@

Figure (a) shows the time and space complexity of tensor network contraction for variety of graph sizes, where the space complexity is the same as the tree width of the problem graph.
The contraction order is obtained using the local search algorithm in Ref.~\cite{Kalachev2021}.
In practice, slicing technique is used for graphs with tree width greater than 27 to fit the computation into a 16GB memory.
One can see all the computing times in figure (b), (c) and (d) have a strong correlation with the tree width.
%The computations shown in \Fig{fig:benchmark} include independent set counting, finding maximum independent set size (panel b),
%computing independence polynomial (panel c) and find one independent set and enumerating independent sets (panel d).
Among these benchmarks, computational tasks with data types \texttt{T (CPU)},
\texttt{$\mathbbm{R}$ (CPU)}, \texttt{$\mathbbm{R}$ (GPU)}, \texttt{$\mathbbm{C}$ (CPU)}, \texttt{$\mathbbm{C}$ (GPU)} and \texttt{T+bounding (CPU)} can utilize fast BLAS functions to evaluate tensor contractions, hence 
are much faster comparing to non-BLAS element types in the same category.
GPU computes much faster than CPU in all cases when the problem scale is large enough such that the actual computing time is comparable or larger than the launching overhead of CUDA kernels.
Most algebras can be computed on GPU, except those requiring dynamic sized structures, i.e. \texttt{PN} and \texttt{SN}.
In figure (c), one can see the Fourier transformation based method is the fastest in computing the independence polynomial,
however, it maybe suffer from the round off errors. The finite field (GF$(p)$) approach is the only method that does not have round-off errors and can run on a GPU.
In figure (d), one can see the technique to bound the enumeration space improves the performance for more than one order in enumerating the MISs.
Bounding can also reduce the memory usage significantly, without which the largest computable graph size is only $\sim150$ on a device with 32GB main memory.

@@

\subsubsection{Number of independent sets and hard-square entropy constant}
We compute the counting of all independent sets for graphs shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The graphs include: the square grid graphs, denoted as SG; the square grid graphs with a filling factor $p$, denoted as SG$(p)$, where $\lfloor pL^{2} \rceil$ square grids are occupied with vertices;
the King's graphs, denoted as K; the King's graphs with a filling factor $p$, denoted as K$(p)$. 
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequences for $L=38$ and $L=39$, which is not known before to our knowledge.
In the computation, we have used the finite field algebra trick for contracting arbitrarily high precision integer tensor networks.
We also computed the quantity $F(L,L)^{1/\lfloor pL^2\rceil}$ as a function of grid size $L$ for these lattices,
where $F(L,L)$ is the number of independent sets of a given grid size.
The results are shown in \Fig{fig:hardsquare}.
Our result for $p=1$ matches those in the literature, and the lattices with hole has much larger average value,
which has a phsical meaning of the hard lattices with holes have much larger mean entropy than those without holes at the high temperature limit.
